---
output: 
  github_document:
    pandoc_args: --webtex=https://latex.codecogs.com/svg.latex?
bibliography: README.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(digits = 4)
```

# Oumuamua

[![Build Status on Travis](https://travis-ci.org/boennecd/Oumuamua.svg?branch=master,osx)](https://travis-ci.org/boennecd/Oumuamua)


The `Oumuamua` contains a parallel implementation of the Multivariate 
Adaptive Regression Splines algorithm suggested in @Friedman91. The package 
can be installed from Github by calling 

```{r install_package, eval = FALSE}
devtools::install_github("boennecd/Oumuamua")
```

There is yet no backronym for the package. 
The rest of this README contains simulation examples where this package 
is compared with the `earth` package. The simulation examples are from
@Friedman91 and are shown to illustrate how to use the package. A discussion
about implementation differences between the `earth` package and this package 
is given at the end. Some comments about future tasks are also listed at the 
end.

## Additive function

We start of with a model with the additive model shown in @Friedman91 [pp. 35].
The model is 

$$y = 0.1 \exp 4 x_1 + \frac 4{1 + \exp-20(x_2 - 1/2)} + 3x_3 + 2x_4 + x_5 + 
   \epsilon$$
   
where $\epsilon$ follows a standard normal distribution and the $x_i$s are 
uniformly distributed on $(0,1)$. Moreover, we let the $x_i$s be correlated 
and introduce a fixed number of noisy correlated variables. 

We start by defining two function to simulate the covariates and outcomes. 

```{r addit_sim}
# generates correlated uniform variables 
get_covs <- function(N, p, rho = .5){
  X <- matrix(rnorm(N * p), nc = p)
  Sig <- diag(p)
  Sig[lower.tri(Sig)] <- Sig[upper.tri(Sig)] <- rho
  X <- X %*% chol(Sig)
  apply(X, 2, pnorm)
}

# simulates data 
additiv_sim <- function(N, p){
  p <- max(5L, p)
  x <- get_covs(N = N, p = p)
  y <- .1 * exp(4 * x[, 1]) + 4 / (1 + exp(-20 * (x[, 2] - 1/2))) + 
    3 * x[, 3] + 2 * x[, 4] + x[, 5] + rnorm(N)
  data.frame(y = y, x)
}
```

Then we run the simulations.

```{r pre_addit_res, message = FALSE}
library(earth)
library(Oumuamua)
```

```{r sim_params, cache = 1}
# parameters in simulation
test_size <- 10000
N <- c(100, 200, 500)
p <- 10
```

```{r addit_performance, cache = 1, dependson = "sim_params"}
# returns minspan and endspan arguments. Similar to suggestion in Friedman 
# (1991) though without adjusting N for number of striclty positive elements 
# in the basis function.
get_spans <- function(N, p, alpha = .05){
  Np <- N * p
  minspan <- as.integer(ceiling(-log2(-1/Np * log1p(-alpha)) / 2.5))
  endspan <- as.integer(ceiling(3 -log2(alpha / p)))
  c(minspan = minspan, endspan = endspan)
}

# functions to fit models
earth_call <- function(sims){
  spans <- get_spans(N = nrow(sims), p = p)
  earth(y ~ ., data = sims, minspan = spans["minspan"], 
        endspan = spans["endspan"], degree = 1, penalty = 2)
}
oumua_call <- function(sims, n_threads = 5L){
  spans <- get_spans(N = nrow(sims), p = p)
  oumua(y ~ ., data = sims, control = oumua.control(
    minspan = spans["minspan"], endspan = spans["endspan"], degree = 1L, 
    penalty = 2, lambda = 1, n_threads = n_threads))
}

# run simulations
set.seed(3779892)
res <- lapply(N, function(N_i){
  # data used for validation
  test_dat <- additiv_sim(test_size, p)
  
  replicate(1000, {
    # simulate
    sims <- additiv_sim(N_i, p)
    
    # fit models
    efit <- earth_call(sims)
    ofit <- oumua_call(sims)
    
    # compute MSE and return
    mse <- function(fit){
      yhat <- predict(fit, newdata = test_dat)
      mean((test_dat$y - yhat)^2)
    }
    
    c(earth = mse(efit), oumua = mse(ofit))
  })
})
```

We make 1000 simulation above for the different sample sizes in the vector 
`N`. We use `p` covariates (`r p`) of which only 5 are associated with the 
outcome. We only allow for an additive model by setting `degree = 1`. The 
`penalty` argument is parameter in the generalized cross validation criteria
mentioned in @Friedman91 [pp. 19-22]. The `lambda` parameter is the L2 penalty 
mentioned in @Friedman91 [pp. 32]. The mean squared errors for each sample 
size is shown below. 

```{r show_addit_performance}
# stats for mean square error
names(res) <- N
lapply(res, function(x) apply(x, 1, function(z) 
  c(mean = mean(z), `standard error` = sd(z) / sqrt(length(z)))))
```

A comparison of computation times with both 1 and 5 threads is given below. 

```{r load_micro_benchmark}
library(microbenchmark)
```

```{r addit_runtimes, cache = 1}
set.seed(17039344)
addi_runtimes <- local({
  run_dat <- additiv_sim(10000, 10)  
  microbenchmark(
    earth = earth_call(run_dat), 
    `oumua (1 thread) ` = oumua_call(run_dat, n_threads = 1L),
    `oumua (5 threads)` = oumua_call(run_dat, n_threads = 5L),
    times = 100)
})
```

```{r show_addit_runtimes}
addi_runtimes
```

## Interaction Example

Next, we consider the non-additive model in @Friedman91 [pp. 37]. The true 
model is 

$$y = \sin\pi x_1x_2 + 20(x_3 - 1/2)^2 + 10x_4 + 5x_5 + \epsilon$$

We define a function to simulate the covariates and outcomes.

```{r def_interact_sim}
interact_sim <- function(N, p){
  p <- max(5L, p)
  x <- get_covs(N = N, p = p)
  y <- 10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 1/2)^2 + 
    10 * x[, 4] + 5 * x[, 5] + rnorm(N)
  data.frame(y = y, x)
}
```

Then we perform the simulation.

```{r interact_sim, cache = 1, dependson = "sim_params"}
# functions to fit models
earth_call <- function(sims){
  spans <- get_spans(N = nrow(sims), p = p)
  earth(y ~ ., data = sims, minspan = spans["minspan"], 
        endspan = spans["endspan"], degree = 3, penalty = 3, nk = 50, 
        fast.k = 0)
}
oumua_call <- function(sims, n_threads = 5L){
  spans <- get_spans(N = nrow(sims), p = p)
  oumua(y ~ ., data = sims, control = oumua.control(
    minspan = spans["minspan"], endspan = spans["endspan"], degree = 3L, 
    penalty = 3, nk = 50L, lambda = 1, n_threads = n_threads))
}

# run simulations
set.seed(3779892)
res <- lapply(N, function(N_i){
  # data used for validation
  test_dat <- interact_sim(test_size, p)
  
  replicate(1000, {
    # simulate
    sims <- interact_sim(N_i, p)
    
    # fit models
    efit <- earth_call(sims)
    ofit <- oumua_call(sims)
    
    # compute MSE and return
    mse <- function(fit){
      yhat <- predict(fit, newdata = test_dat)
      mean((test_dat$y - yhat)^2)
    }
    
    c(earth = mse(efit), oumua = mse(ofit))
  })
})
```

We have increased `degree` to allow for interactions. We also increase 
`penalty` as suggested in @Friedman91 (though some further tuning might be 
needed). We set `fast.k = 0` in the `earth` call as this feature is not yet
implemented in this package. The mean squared error estimate is shown below. 

```{r show_interact_sim}
# stats for mean square error
names(res) <- N
lapply(res, function(x) apply(x, 1, function(z) 
  c(mean = mean(z), `standard error` = sd(z) / sqrt(length(z)))))
```

A comparison of computation times with both 1 and 5 threads is given below.

```{r interact_runtimes, cache = 1, dependson = "interact_sim"}
set.seed(17039344)
inter_runtimes <- local({
  run_dat <- interact_sim(10000, 10)  
  microbenchmark(
    earth = earth_call(run_dat), 
    `oumua (1 thread) ` = oumua_call(run_dat, n_threads = 1L),
    `oumua (5 threads)` = oumua_call(run_dat, n_threads = 5L),
    times = 10)
})
```

```{r show_interact_runtimes}
inter_runtimes
```

## Interaction Example with Factor 
We add a dummy variable to the model from before in the last example. The code 
is very similar.

```{r def_dum}
factor_sim <- function(N, p){
  p <- max(5L, p)
  x <- get_covs(N = N, p = p)
  
  n_grp <- 5L
  fac <- gl(n_grp, N / n_grp)
  grp_effect <- seq(-3, 3, length.out = n_grp) 
    
  y <- 10 * sin(pi * x[, 1] * x[, 2]) + 20 * (x[, 3] - 1/2)^2 + 
    10 * x[, 4] + 5 * x[, 5] + grp_effect[as.integer(fac)] + rnorm(N)
  data.frame(y = y, x, fac = fac)
}
```

```{r dum_sim, cache = 1, dependson = "sim_params"}
# functions to fit models
earth_call <- function(sims){
  spans <- get_spans(N = nrow(sims), p = p)
  earth(y ~ ., data = sims, minspan = spans["minspan"], 
        endspan = spans["endspan"], degree = 3, penalty = 3, nk = 50, 
        fast.k = 0, fast.beta = 0)
}
oumua_call <- function(sims, lambda = 1, n_threads = 5L){
  spans <- get_spans(N = nrow(sims), p = p)
  oumua(y ~ ., data = sims, control = oumua.control(
    minspan = spans["minspan"], endspan = spans["endspan"], degree = 3L, 
    penalty = 3, lambda = lambda, nk = 50L, n_threads = n_threads))
}

# run simulations
set.seed(3779892)
res <- lapply(N, function(N_i){
  # data used for validation
  test_dat <- factor_sim(test_size, p)
  
  replicate(1000, {
    # simulate
    sims <- factor_sim(N_i, p)
    
    # fit models
    efit <- earth_call(sims)
    ofit <- oumua_call(sims)
    
    # compute MSE and return
    mse <- function(fit){
      yhat <- predict(fit, newdata = test_dat)
      mean((test_dat$y - yhat)^2)
    }
    
    c(earth = mse(efit), oumua = mse(ofit))
  })
})
```

The mean squared errors are given below.

```{r show_dum_sim}
# stats for mean square error
names(res) <- N
lapply(res, function(x) apply(x, 1, function(z) 
  c(mean = mean(z), `standard error` = sd(z) / sqrt(length(z)))))
```

A comparison of computation times with both 1 and 5 threads is given below.

```{r dum_runtimes, cache = 1, dependson = "dum_sim"}
set.seed(17039344)
factor_runtimes <- local({
  run_dat <- factor_sim(10000, 10)  
  microbenchmark(
    earth = earth_call(run_dat), 
    `oumua (1 thread) ` = oumua_call(run_dat, n_threads = 1L),
    `oumua (5 threads)` = oumua_call(run_dat, n_threads = 5L),
    times = 10)
})
```

```{r show_dum_runtimes}
factor_runtimes
```

## Interaction Example (deep)

```{r def_deep_interact_sim}
interact_sim <- function(N, p){
  p <- max(5L, p)
  x <- get_covs(N = N, p = p)
  y <- sin(pi * (x[, 1] + x[, 2] + x[, 3]         )) + 
       sin(pi * (         x[, 2] + x[, 3] + x[, 4])) + 
       rnorm(N)
  data.frame(y = y, x)
}
```

```{r deep_sim, cache = 1}
earth_call <- function(sims){
  spans <- get_spans(N = nrow(sims), p = p)
  earth(y ~ ., data = sims, minspan = spans["minspan"], 
        endspan = spans["endspan"], degree = 3, penalty = 3, nk = 100, 
        fast.k = 0, fast.beta = 0)
}
oumua_call <- function(sims, lambda = 1, n_threads = 5L){
  spans <- get_spans(N = nrow(sims), p = p)
  oumua(y ~ ., data = sims, control = oumua.control(
    minspan = spans["minspan"], endspan = spans["endspan"], degree = 3L, 
    penalty = 3, lambda = lambda, nk = 100L, n_threads = n_threads))
}

# run simulations
set.seed(3779892)
res <- lapply(10000, function(N_i){
  # data used for validation
  test_dat <- interact_sim(test_size, p)
  
  replicate(100, {
    # simulate
    sims <- interact_sim(N_i, p)
    
    # fit models
    efit <- earth_call(sims)
    ofit <- oumua_call(sims)
    
    # compute MSE and return
    mse <- function(fit){
      yhat <- predict(fit, newdata = test_dat)
      mean((test_dat$y - yhat)^2)
    }
    
    c(earth = mse(efit), oumua = mse(ofit))
  })
})
```

The mean squared errors are given below.

```{r show_deep_sim}
# stats for mean square error
lapply(res, function(x) apply(x, 1, function(z) 
  c(mean = mean(z), `standard error` = sd(z) / sqrt(length(z)))))
```

A comparison of computation times with both 1 and 5 threads is given below.

```{r deep_runtimes, cache = 1, dependson = "deep_sim"}
set.seed(17039344)
deep_runtimes <- local({
  run_dat <- interact_sim(10000, 10)  
  microbenchmark(
    earth = earth_call(run_dat), 
    `oumua (1 thread) ` = oumua_call(run_dat, n_threads = 1L),
    `oumua (5 threads)` = oumua_call(run_dat, n_threads = 5L),
    times = 10)
})
```

```{r show_dum_runtimes}
deep_runtimes
```


## Comparison with the earth Package

Some of the main differences between this package and the `earth` package is 

 - the `earth` package has many more features!
 - the `earth` package is single threaded (as of this writing).
 - the `earth` includes the suggestions in @Friedman93 to decrease the computation time. It would be nice to include these in this package. It only matters when there are many basis functions.
 - the `earth` package does not include an L2 penalty which simplifies  the computation of the generalized cross validation criterion at each knot. 
 - the `earth` package creates an orthogonal design matrix during the 
   estimation which allows one (I think?) to skip some computations of the 
   generalized cross validation criterion (a "full" forward and backward
   substitution). Put differently, there are no forward or backward 
   substitutions in the C function that finds the knots in the `earth` 
   package.

Some computation can be skipped if one sets `lambda` to zero (i.e., no L2 
penalty). The following code blocks shows the impact. 

```{r zero_lambda_dum_runtimes, cache = 1, dependson = "dum_sim"}
set.seed(17039344)
factor_runtimes <- local({
  run_dat <- factor_sim(10000, 10)  
  microbenchmark(
    earth = earth_call(run_dat), 
    `oumua (1 thread) ` = oumua_call(run_dat, n_threads = 1L, lambda = 0),
    `oumua (5 threads)` = oumua_call(run_dat, n_threads = 5L, lambda = 0),
    times = 10)
})
```

```{r zero_show_dum_runtimes}
factor_runtimes
```

Settings `lambda = 0` yields one less back substitution for each knot position. 
However, this is not preferred as the implementation is not numerical stable 
in some cases.

A final issue that still needs to be addressed is the L2 penalty in this 
package. When the knot position is found then the L2 penalty is applied to the 
coefficients in the model

$$y = \dots + \beta_1x_i + \beta_2(x_i - k)_+$$

where $k$ is the knot and $\cdots$ are the other terms already included in 
the model. 
However, the final L2 penalty is applied to the 
coefficients in the model 

$$y = \dots + \beta_1(k - x_i)_+ + \beta_2(x_i - k)_+$$

which does not yield an equivalent model. The former is faster during the knot 
estimation while the latter is faster later as it yields a more sparse design 
matrix and thus faster computation times later.

## References
